\chapter{Tangent Assembly: Inactive and Alternative Paths}
\label{app:tangent}

This appendix chapter collects tangent paths that remain in the repository but
are not part of the default parallel PMG benchmark.

\section{Global MATLAB-Style Reference Path}

\code{build\_global\_tangent\_matrix(...)} still assembles the full global
matrix directly as
\[
K = B^T D_S B.
\]

This is the shortest conceptual bridge to MATLAB and remains valuable for
serial debugging and parity checks, but it is not the normal PMG-path solve.

\section{Unique-Owner Constitutive Recovery}

The owned tangent pattern can also be prepared with unique-owner support:
\begin{itemize}
\item \code{include\_unique = true},
\item optional gather or neighbor exchange metadata,
\item mapping arrays that recover overlap-local constitutive data from unique
owners.
\end{itemize}

Those structures matter for \code{unique}, \code{unique\_gather}, and
\code{unique\_exchange} modes. They are inactive in the committed benchmark
because \code{constitutive\_mode = overlap}.

\section{Legacy Scatter Kernel}

The \code{legacy} kernel is the closer relative of the MATLAB assembly mental
model:
\begin{itemize}
\item build an element matrix,
\item flatten it,
\item scatter entries through a precomputed map.
\end{itemize}

It is still useful as a reference and validation kernel, but it is not the
default production kernel for the 3D PMG benchmark.

\section{BDDC Local-Subdomain Pattern}

\code{BDDCSubdomainPattern} serves a different PETSc consumer: local square
subdomain matrices for MATIS/PCBDDC. Its main differences from the owned-row
pattern are:
\begin{itemize}
\item local square numbering instead of global owned rows,
\item explicit interface and adjacency metadata,
\item tangent refresh routines specialized for the BDDC preconditioner.
\end{itemize}

This branch becomes relevant only when the solver backend is changed to
\code{bddc}.

\section{Why The Main Text Excludes Them}

The mainline chapter now follows one concrete path:
\[
\text{owned rows} \rightarrow \text{overlap constitutive data} \rightarrow
\text{rows kernel}.
\]

The inactive paths are kept here so they remain documented without obscuring
which branch the default benchmark actually executes.
