\chapter{Linear Solvers and PETSc: Inactive Backends and Policies}
\label{app:linear}

This appendix chapter collects solver families, preconditioner backends, and
policy knobs that exist in the repository but are not part of the committed
parallel PMG benchmark.

\section{Other Solver Families}

The factory in \petscmod{linear/solver.py} also supports:
\begin{itemize}
\item direct solvers,
\item pure Python deflated FGMRES,
\item PETSc Krylov solves without the MATLAB-style wrapper,
\item experimental PETSc-native deflation variants.
\end{itemize}

These are useful for debugging, comparison, or research, but not the default
benchmark path.

\section{Other Backends}

The same outer solver family can run with several backends:
\begin{itemize}
\item \code{hypre},
\item \code{gamg},
\item \code{bddc},
\item built-in PETSc \code{pmg},
\item LU/direct factorization.
\end{itemize}

Only \code{pmg\_shell} is mainline in the committed 3D benchmark.

\section{Why BDDC Is Separate}

The BDDC backend changes more than a PETSc option. It changes the matrix shape
needed by the preconditioner, because it wants the local subdomain operators
built through \code{BDDCSubdomainPattern}. That is why BDDC is kept out of the
mainline text.

\section{Other Explicit Policy Controls}

The broader solver framework exposes policies that the PMG-shell mainline path
does not vary:
\begin{itemize}
\item \code{preconditioner\_matrix\_source} other than \code{tangent},
\item lagged or accepted-step matrix policies,
\item rebuild intervals other than every Newton iteration,
\item full-system preconditioning modes.
\end{itemize}

Those options remain relevant for alternative backends and experiments.
