\chapter{Newton Solvers and Damping: Inactive Branches}
\label{app:newton}

This appendix chapter collects Newton interfaces and solver interactions that
remain implemented but are not exercised by the committed default PMG
benchmark.

\section{Full-System Operator Mode}

Some solver backends prefer the full operator instead of the reduced free-space
matrix. In those cases Newton can keep a full PETSc operator and let the solver
manage free indices internally. This branch is inactive for \code{pmg\_shell}
because \code{full\_system\_preconditioner = false}.

\section{Local Full-Operator Builders}

The constitutive layer also exposes:
\begin{itemize}
\item \code{build\_F\_local(...)},
\item \code{build\_F\_free\_local(...)},
\item related local-output interfaces.
\end{itemize}

These support PETSc full-operator workflows that consume owned-row local data
directly. They are not the mainline PMG-shell path.

\section{Separate Explicit Preconditioning Matrices}

The broader PETSc framework can request an explicit preconditioning matrix
separate from the operator used in the Krylov solve. Depending on solver policy,
that matrix can come from:
\begin{itemize}
\item the elastic operator,
\item the tangent operator,
\item the regularized operator.
\end{itemize}

The default PMG-shell benchmark keeps a narrower, fixed policy and therefore
does not need the full generality discussed above.

\section{Optional Warm Starts}

The code also supports temporary first-iteration warm-start bases passed from
continuation into Newton. Those features remain inactive in the committed
benchmark because \code{continuation\_first\_newton\_warm\_start\_mode = none}.
